Type theory in type theory with single substitutions

### 2025-01-24 cpp/p.tex
stuff

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck other equations. All the other equations are substitution
(naturality) rules or computation rules for different type
formers. SSC provides a simple, minimalistic alternative to parallel
substitution calculi or B-systems for defining type theory. SSC
relates to CwF as extensional combinatory calculus relates to lambda
calculus. If we have some additional type formers, we show that SSC models
are actually equivalent to CwF-based models in a weak sense. Most
results in this paper were formalised in Agda.
\end{abstract}

@@

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
  only consider well-formed, well-scoped and well-typed abstract
  syntax trees, there are no meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
Because types are isomorphic to certain terms (witnessed by $\El$ and $\cd$),
we have to index types by their (universe) level, and types at lower
levels are included at upper levels by $\Lift$. We use implicit
arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
second-order functions. We say that $\Pi$ binds a $\Tm$-variable
in its second $\Ty$-argument. Every binder is represented by
a second-order function, and as a result, we don't need to talk about
contexts and substitutions when describing the theory: we re-use the
binding features of the metatheory to describe those of our object
theory (this technique is also known as higher-order abstract syntax
\cite{DBLP:conf/lics/Hofmann99}, logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} or
two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
second-order models, thus the semantics of a SOGAT is given by a first
order generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). GATs are generalisations of
single-sorted algebraic theories (such as monoids, groups) to multiple
sorts where later sorts can be indexed over previous sorts. The usual
results from universal algebra (e.g.\ models form a category with
finite limits and initial object, existence of free and co-free
models) carry over to GATs \cite{andras,DBLP:phd/hal/Moeneclaey22}. An
example of a GAT is the theory of categories where there is a sort of
objects and a sort of morphisms double-indexed over objects.

@@

The single substitution semantics is in some sense too minimalistic:
for example, the equation $b[\p^+][\langle\q\rangle] = b$ is not
\emph{derivable} in any model, but it is \emph{admissible} in the
syntax. (Here $b : \Tm\,(\Gamma\ext A)\,B$ meaning that it is a term
which depends on a context $\Gamma$ and an extra variable of type $A$;
$b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p^+])$ is a version
where we weakened \emph{before} the last variable; in
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ we substituted
the last variable for the previous one.)  This is analogous to
e.g.\ parametricity results \cite{DBLP:journals/jfp/BernardyJP12},
which do not hold in an arbitrary model, but they hold in the
syntax. In this paper, we show that for the particular SOGAT
(\ref{eq:tt}), the \emph{syntaxes} of the single and parallel
semantics are isomorphic. The techniques that we use for this proof
are not specific to the SOGAT (\ref{eq:tt}), we expect
that they work for any SOGAT. Isomorphism of the syntaxes means that
the corresponding sorts ($\Con$, $\Ty$, $\Tm$) are isomorphic. For a
general SOGAT, there are more models of the correponding single
substitution GAT than the parallel substitution GAT. Every parallel
model is also a single model, but not necessarily the other way. While
B-systems are equivalent to CwFs (and C-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more single
substitution models of (\ref{eq:tytm}) than parallel substitution
models (CwFs). The situation is similar to that of extensional combinatory
calculus and lambda calculus, where the syntaxes are equivalent, and
every lambda model is a combinatory model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
relationship of monoids over a set $X$ and nil/cons algebras over $X$,
where every monoid is a nil/cons-algebra, but not the other way; it is
well-known that free monoids over $X$ (syntax of monoid over $X$) are
the same as lists of $X$ (the syntax of nil/cons algebras over $X$).

@@

Theory (\ref{eq:tt}) can be extended
with unit and $\Sigma$ types, more precisely with the following rules.
\begin{equation}\label{eq:sigma}
  \begin{alignedat}{10}
    & \top && : \Ty\,0 \\
    & \tt && : \Tm\,\top \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
\end{alignedat}
\end{equation}
Using the same insight about types and the universe, the fact that
dependences can be represented by function spaces, and that
collections of dependencies can be collected using $\Sigma$ types, we
can actually derive a parallel model from any single substitution
model. This way we show that for the particular theory (\ref{eq:tt})+(\ref{eq:sigma}),
single and parallel substitution models are actually equivalent in a
weak sense.

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Sections \ref{sec:tt} and
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda, we use a strict Prop-valued
\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
coercion rule (transport rule) and postulated QIITs with computation
rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
uniqueness of identity proofs (UIP), and our metatheory is
incompatible with homotopy type theory \cite{HoTTbook}. Our notation
on paper is close to Agda's, we write $\Pi$ types as $(x:A)\ra B$,
$\Sigma$ types as $(x:A)\times B$, and use implicit arguments and
overloading extensively, for example $\blank[\blank]$ for types and
terms overloaded.

@@

>>>>

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck the other equations. SSC provides a simple, minimalistic
alternative to parallel substitution calculi or B-systems for defining
type theory. SSC relates to CwF as extensional combinatory calculus
relates to lambda calculus: there are more models of the former, but
the syntaxes are equivalent. If we have some additional type formers,
we show that SSC models are actually equivalent to CwF-based models in
a weak sense. Most results in this paper were formalised in Agda.
\end{abstract}

@@

In this paper, we will answer this question by defining type theory as
a generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). This has the following
properties.
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic
  \cite{alti:phd93,abel2013normalization,DBLP:conf/cpp/AdjedjLMPP24}. That
  is, we only consider well-formed, well-scoped and well-typed
  abstract syntax trees, there are no meaningless terms.
\item[(ii)] The equations of the algebraic theory describe the
  (symmetric) conversion relation.
\end{enumerate}
For type theory, such an intrinsic quotiented presentation is very
natural as typing depends on conversion. The algebraic approach avoids
lots of boilerplate coming from using an explicit conversion relation
(as in e.g.\ \cite{chapman09eatitself}). In the algebraic approach,
syntax and semantics are concordant: the syntax is simply the initial
model, which always exists for any GAT
\cite{DBLP:journals/pacmpl/KaposiKA19}. Such a syntax is quotiented
which means that any function defined on the syntax needs to respect
conversion. For example, it is not possible to define a printing
function which distinguishes the terms $(\lambda x.x)\,1$ and $1$, as
these are convertible. However, it is still possible to define
normalisation \cite{lmcs:4005} and typechecking \cite{typecheck} at
this level of abstraction.

The algebraic/intrinsic approach is also more abstract than extrinsic
approaches, which means that we don't have to make ad-hoc choices. For
example, we do not have to decide whether we want function space {\`a}
la Curry or Church, whether we work with paranoid or economic rules
\cite[Section 5]{winterhalter:hal-01626651}, or whether we have
explicit or recursively defined substitution (instantiation of
variables by terms). These choices are forced by the notion of GAT:
the domain of a lambda appears in its type, se we are always {\`a} la
Church; algebraic theories don't allow recursively defined operators,
so instantiation is an explicit operation, and so on. However, there
are still some options for the exact definition of the substitution
calculus: most notions of type theory use parallel substitutions,
which means that the instantiation operation replaces all variables in
a context by terms in another context \emph{at the same time}. Some
examples of algebraic definitions of type theory using parallel
substitutions ranging from the more syntactic to the more semantic:
Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories with
families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}), Awodey's
natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
\cite{10.1145/3204492}. In traditional extrinsic presentations
intantiation replaces one variable by a single term. To bridge the gap
between the intrinsic parallel substitution world and the extrinsic
single substitution world, Voevodsky introduced B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
desription of a single substitution calculus. B-systems involve
complex rules describing telescopes, weakenings and substitutions
under telescopes, with several equations.

In this paper, we introduce a new single substitution calculus which
is simpler and more minimalistic than the previous algebraic
approaches. It shows that to explain type theory in an algebraic way,
there is no need for categories, parallel substitutions or weakenings,
empty substitution, telescopes, or combinations of these. All our
operations are easy to motivate, we showcase this in Section
\ref{sec:tt} which is a tutorial introduction to our theory. Our
single substitution calculus is in a sweet spot: our notations read
like traditional extrinsic syntactic presentations, but they are
algebraic, thus come with a well behaved model theory, and can be
easily related to more semantic descriptions.

Our single substitution theory is in some sense too minimalistic: for
example, the equation $b[\p^+][\langle\q\rangle] = b$ is not
\emph{derivable} in any model, but it is \emph{admissible} in the
syntax. (Here $b : \Tm\,(\Gamma\ext A)\,B$ is a term which depends on
a context $\Gamma$ and an extra variable of type $A$; $b[\p^+] :
\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p^+])$ is a version where we
weakened \emph{before} the last variable; in
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ we substituted
the last variable for the previous one.)  This is analogous to
e.g.\ parametricity results \cite{DBLP:journals/jfp/BernardyJP12},
which do not hold in an arbitrary model, but they hold in the
syntax. In this paper, we show that the \emph{syntax} of our single
substitution theory and the syntax of the corresponding parallel
substitution theory (CwF-based theory) are isomorphic. All CwF-based
models are models of our single substitution theory, but not
necessarily the other way. While B-systems are equivalent to CwFs (and
C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
single substitution models than CwF-based models. This relationship is
similar to that of extensional combinatory calculus and lambda
calculus, where the syntaxes are equivalent, and every lambda model is
a combinatory model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
relationship between monoids over a set $X$ and nil/cons algebras over
$X$, where every monoid is a nil/cons-algebra, but not the other way;
but it is well-known that free monoids over $X$ (syntax of monoid over
$X$) are the same as lists of $X$ (the syntax of nil/cons algebras
over $X$).

Technology to show that syntaxes are equivalent: alpha normalisation
-- terms which don't include instantiation, but are still quotiented.

Minimisation: (i) if we have a universe representing every type as a
term, we can remove some equation, obtaining a more minimalistic, but
isomorphic GAT. (ii) if we even have Sigma tyes, we can actually
derive a parallel model from an SSC model.

SOGAT stuff

The techniques that we use for this proof are not specific to the
SOGAT (\ref{eq:tt}), we expect that they work for any
SOGAT. Isomorphism of the syntaxes means that the corresponding sorts
($\Con$, $\Ty$, $\Tm$) are isomorphic. For a general SOGAT, there are
more models of the corresponding single substitution GAT than the
parallel substitution GAT.

xxx

For formalisation on a computer, the algebraic approach also has its
downsides: as mutually defined recursive operations are not allowed,
substitution (instantiation of variables by terms) is explicit, which
means that it is a constructor, and does not compute by
definition. For example, $(t\,a)[x\mapsto u] = (t[x\mapsto
  u])\,(a[x\mapsto u])$ is a derivable equation in the syntax, but it
is not definitional. This makes formalisation difficult, and results
in the situation called ``transport hell''.

Related work:
- SOGAT->GAT conversion
- B systems
- others

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:

For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
Because types are isomorphic to certain terms (witnessed by $\El$ and $\cd$),
we have to index types by their (universe) level, and types at lower
levels are included at upper levels by $\Lift$. We use implicit
arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
second-order functions. We say that $\Pi$ binds a $\Tm$-variable
in its second $\Ty$-argument. Every binder is represented by
a second-order function, and as a result, we don't need to talk about
contexts and substitutions when describing the theory: we re-use the
binding features of the metatheory to describe those of our object
theory (this technique is also known as higher-order abstract syntax
\cite{DBLP:conf/lics/Hofmann99}, logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} or
two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
second-order models, thus the semantics of a SOGAT is given by a first
order generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). GATs are generalisations of
single-sorted algebraic theories (such as monoids, groups) to multiple
sorts where later sorts can be indexed over previous sorts. The usual
results from universal algebra (e.g.\ models form a category with
finite limits and initial object, existence of free and co-free
models) carry over to GATs \cite{andras,DBLP:phd/hal/Moeneclaey22}. An
example of a GAT is the theory of categories where there is a sort of
objects and a sort of morphisms double-indexed over objects.

@@

@@

Theory (\ref{eq:tt}) can be extended
with unit and $\Sigma$ types, more precisely with the following rules.
\begin{equation}\label{eq:sigma}
  \begin{alignedat}{10}
    & \top && : \Ty\,0 \\
    & \tt && : \Tm\,\top \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
\end{alignedat}
\end{equation}
Using the same insight about types and the universe, the fact that
dependencies can be represented by function spaces, and that
collections of dependencies can be collected using $\Sigma$ types, we
can actually derive a parallel model from any single substitution
model. This way we show that for the particular theory (\ref{eq:tt})+(\ref{eq:sigma}),
single and parallel substitution models are actually equivalent in a
weak sense.

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Sections \ref{sec:tt} and
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda, we use a strict Prop-valued
\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
coercion rule (transport rule) and postulated QIITs with computation
rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
uniqueness of identity proofs (UIP), and our metatheory is
incompatible with homotopy type theory \cite{HoTTbook}. Our notation
on paper is close to Agda's: we write $\Pi$ types as $(x:A)\ra B$,
$\Sigma$ types as $(x:A)\times B$, we use implicit arguments and
overloading extensively, for example instantiation $\blank[\blank]$ is
overloaded for types and terms.

@@

TODO: named CwFs would be even closer to usual syntax.
